---
title: "DB_Object Demo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("..\\code\\pacala_db.R")
```

```{r, echo=F}
c_file_path = 'C:\\dev\\pacala\\data'
c_data_file_name = 'pacala_data.txt'
c_params_file_name = 'pacala_parameters.txt'
```

### Motivation
* The data format changes sometimes, and is spread over a number of files.
* Be nice to have a schema driven data layer that abstracts the implementation and allows simplified interaction with the data.

# Source Data
Starting with data that look like this
```{r}
data_in_file = load_pacala_data(c_file_path, c_data_file_name)
```
```{r, echo=F}
data_in_file
```

and paremters like this
```{r}
params_in_file = load_pacala_params(c_file_path, c_params_file_name)
```
```{r, echo=F}
params_in_file
```

# Data Structure

### Set up
Here the schema is loaded
```{r}
  schema_list = load_schema()  
```
```{r, echo=F}
  schema_list
```
and used to set up the data object. The internal schema is shown
```{r}
  db_obj = setup_db(schema_list)
  db_obj[['state']]
```

### Inserts & Retrieves
Here the data loaded from the file is inserted. The internal db is shown
```{r}
  db_obj = insert_data_into_db(db_obj, data_in_file)
  db_obj[['data']]
```

Here a dataframe is retrieved with a single command, its trimmed to populted columns 
```{r}
curr_db = get_current_data(db_obj)
```
```{r, echo=F}
curr_db
```

The parameters are inserted
```{r}
  db_obj = insert_db_params(db_obj, params_in_file)
```
```{r, echo=F}
  db_obj[['data']]
```

and now the same single command returns a larger dataframe
```{r}
curr_db = get_current_data(db_obj)
```
```{r, echo=F}
curr_db
```

Can also executed canned queries, here looking up a species by its ID
```{r}
lookup_name = get_species_of_individ(db_obj, 2)
```
```{r, echo=F}
lookup_name
```

# Graphics
```{r}
draw_landscape(curr_db, 1000, 1000)
```