---
title: " Pacala Forest Simulator"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('C:\\dev\\pacala\\code\\pacala.R')
source('C:\\dev\\pacala\\code\\pacala_db.R')
```

Pacala, Canham, Silander (1993) Can J For Res 23: 1980-88

###Objective
* spatailly explicit model of forest dynamic
* based on submodel predicting growth, dispersal, survival, recruitment
* based on submodel predicting local availability of resources
* competition is purely mechanistic, only interact by resource depletion

```{r}
c.x.max = 100
c.y.max = 100
c_file_path = 'C:\\dev\\pacala\\data'
c_data_file_name = 'pacala_data.txt'
c_params_file_name = 'pacala_parameters.txt'

```

###Model design
Using a $c.x.max$ x $c.y.max$ grid 
With circular, non-overlapping trees

## Exploring the data
Load the data and look at it
```{r load model data, echo=F}
data_to_insert = load_pacala_data(c_file_path, c_data_file_name)
params_to_insert = load_pacala_params(c_file_path, c_params_file_name)
db_obj = setup_db_obj(load_schema())
```
```{r, echo=F}
print (data_to_insert)
print (params_to_insert)
print(db_obj[['state']][['species_levels']])
```


## The data is loaded into the data structure
```{r}
db_obj = insert_individuals(db_obj, data_to_insert)
db_obj = insert_species_params(db_obj, params_to_insert)
```
```{r, echo=F}
print(get_current_data(db_obj))
```

### Plot the data
```{r}
draw_landscape(db_obj, 10, 10)
```

### Generate resource availibity for each individual
using a light model of a single point source at origin with any point's intesity proportional to it distance from the source
```{r}
db_obj = run_resources_submodel(db_obj)
```
```{r, echo=F}
get_derived_values(db_obj, c('light'))
```

### Generate growth for each individual
```{r}
db_obj = run_growth_submodel(db_obj)
```
```{r, echo=F}
get_derived_values(db_obj, c('ring_width', 'tree_height'))

```

### Generate mortality probability for each individual
```{r}
db_obj = run_mortality_submodel(db_obj)
```
```{r, echo=F}
get_derived_values(db_obj, c('growth_dist', 'mortality_func', 'Y_L', 'Y_D'))
```

### Generate recruitment surface for each individual
```{r, eval=F}
db_obj = run_recruitment_submodel(db_obj, 100, 100)
```
```{r, echo=F, eval=F}
for (i in 1:length(db_obj[['surfaces']])) {
  heatmap(db_obj[['surfaces']][[i]], Rowv = NA, Colv = "Rowv")
}  
```



