---
title: "Literate Programming & Documentation"
subtitle: "RRROBOTS Week 3"
author: "Hao Ye"
date: "2017, April 18"
output:
  xaringan::moon_reader:
    includes:
      after_body: scripts.html
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
    
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
# Lesson Objectives

### Project Status
### Literate Programming
### Documentation

---
class: center, middle, inverse

# Project Status

---

# Project Check-in

### What stage are you at?
* Paper / group defined?

### Develop a Roadmap
* What are your objectives?
* How do you know your project is complete?

---

# GitHub issues

### Remember, GitHub originally intended as a tool for software development!

--

### Why?
* plan out what you are going to do ("issue" == "bug" == "feature")
* be specific about tasks
* organization of work

### Collaborations
* threaded (opt-in) discussion
* prevent duplication of effort and unnecessary emails

---

# Intro to Markdown

### Why learn yet another syntax?
* It allows you to format your docs:
  - GitHub will automatically render `readme` files within folders.
  - If it has a `.md` extension, Markdown formatting will be applied.
* Many sites use Markdown formatting:
  - StackOverflow Qs and Responses
  - GitHub project interfaces
* It has very little added complexity - the raw text is human-readable.
* GitHub Integration: 
  - keywords will automagically close issues:
    https://help.github.com/articles/closing-issues-via-commit-messages/

---

# Basic Markdown

```
# Header level 1
## Header level 2

*italicized text*

**bolded text**

[link text](link URL)

* Bullet Points
  - indented bullets
    1. numbered
    2. list
    3. items
```
---

class: center, middle, inverse

# Literate Programming

---

# What is it?

1. Describe what you want to do in normal language (e.g. English)

2. Write code to do your computations.

  - Hide the messy bits that are not relevant for describing the analysis.
  
3. Generate the output.
  
4. Glue it all together:

  - the story
  - the code
  - the results
  
--

### Yeah, it's sort of like a scientific paper!

---

class: center, middle

# What Can We Use It For?

---

# Use Cases

### Package Vignettes and Tutorials
* https://github.com/tidyverse/ggplot2
* https://github.com/datacarpentry/R-ecology-lesson

### Reports, Supplements, Papers
* https://github.com/SimonGoring/GoringetalPollenRichness
* http://www.collabra.org/article/10.1525/collabra.71/

---

# Advanced Use Cases

* Slides
  - Beamer Template (LaTeX)
  - `xaringan` package (what this is created with)

* Websites
  - `blogdown` package

* Books
  - `bookdown` package
  
* Theses
  - `thesisdown` package

---

# Why Use It?

### No more copy-pasting of numerical results, tables, etc.

### No more toggling back and forth between the text and the code.

--

### It keeps everything organized and consistent!

---
class: center, middle

# Demo

---

# Real-Time Embedding of Code and Output

```{r, message = FALSE, echo = FALSE}
library(ggplot2)
library(plotly)
```

```{r, fig.height = 4, fig.width = 7}
ggplotly(ggplot(iris, aes(x = Petal.Length, y = Sepal.Width, color = Species)) + 
             geom_point() + theme_bw())
```

---
class: center, middle, inverse

# Documentation

---

# Function Documentation

### Goals
* describe what the function does without having to read the code
* clarify the inputs and outputs so that others can build on either end

### Guidelines
1. what it does;
2. what all its inputs are (in order);
3. what it requires of the inputs and the state of the system ("presumes")'
4. what side-effects it may have (e.g., "plots histogram of residuals");
5. what all its outputs are (in order)

---

# Project Documentation

### Map the location of things in the Readme

### Describe the general flow of analysis in scripts or markdown
* Usually I have both
  - a script for testing and running interactively
  - a markdown report of the whole "pipeline" and results
  
### Use GitHub to track stuff!

---

# Package Documentation

R uses packages to extend the language functionality.

### Packages are easy to write
* very little overhead if for personal use
* RStudio will take care of lots of things for you
* uploading to CRAN makes it available for others, requires more extensive documentation
* can be code or data

### Roxygen 
* a package and plug-in for RStudio
* generated `.rd` files, correctly formatted documentation for R packages
* built from specifically formatted comments

---
class: center, middle

# Demo

---

# References

1. [Markdown Guide](https://guides.github.com/features/mastering-markdown/)
2. [Reproducible Manuscripts using RMarkdown](https://elifesciences.org/elife-news/composing-reproducible-manuscripts-using-r-markdown)
3. [Advice on Programming](http://bactra.org/weblog/593.html)

---

class: center, middle

# Acknowledgments

Class supported by:

<img width=50% src = "mozilla_science_logo.png">
<img width=50% src = "open_data_science_logo.png">

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).