---
title: "Power Simulation Notes"
author: "Eiren Jacobson"
date: "April 18, 2017"
output: pdf_document
---

header-includes:
   - \usepackage{caption}
output: 
  pdf_document:
    latex_engine: xelatex

---
\begin{center}
\Large{Eiren Jacobson} 
\end{center}

\begin{center}
\large{`r format(Sys.time(), '%d %B %Y')`}
\end{center}

```{r ChunkOpts, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, results="hide", message=FALSE, warning=FALSE)
```

# Variability in the 3-year passive acoustic dataset

* The mean PPS values measured across all stations and years is shown in the following table

```{r, results="show", echo=FALSE}
 
 require(data.table)
 require(ggplot2)
 require(reshape2)
 require(nlme)
 
 load("~/Desktop/acSim/Data/CpodPPS.RData")
 
 means <-  PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(YEAR, MOORING)]
 mean.table <-  dcast(means, MOORING~YEAR)

 
 mean.table$ALL <- PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(MOORING)]$Mean

d <- PPS.Data[,.(Mean = mean(PPS), Var = var(PPS)), 
                           by = .(MOORING)]

mean.table <- rbind.data.frame(mean.table, 
                               cbind("MOORING"="ALL", "2013"=221.0256, "2014"=451.5736, "2015"=365.3147, "ALL" = 350.781))

mean.table[,2] <- as.numeric(mean.table[,2])
mean.table[,3] <- as.numeric(mean.table[,3])
mean.table[,4] <- as.numeric(mean.table[,4])
mean.table[,5] <- as.numeric(mean.table[,5])

mean.table[,2:5] <- round(mean.table[,2:5], digits=0)

mean.table

```

* I first constructed a linear mixed-effects model where year is a fixed effect and mooring is a random effect.  The response variable is the log(mean PPS) for each mooring and year.


```{r, echo=TRUE, results="markup"}

#names(CXY)[1] <- "MOORING"

#means <- merge(CXY[,c(1,4)], means, by="MOORING")
 
#model <- lme(fixed = log(Mean) ~ as.numeric(factor(YEAR)), 
#             random = ~ 1 | MOORING, data = means)

model <- glm(log(Mean) ~ as.numeric(factor(YEAR)), data=means)

summary(model)

# MODEL CHECKS RUN (RESULTS NOT SHOWN)
# plot(model)
# plot(model, YEAR ~ fitted(.))
# plot(model, MOORING ~ fitted(.))
# plot(model, resid(.) ~ fitted(.))
# qqnorm(model, ~ resid(.) | YEAR)
# qqnorm(model, ~ resid(.) | MOORING)
# qqnorm(model, ~ resid(.) | Density)

```


```{r, echo=FALSE}

plot(means$Mean, exp(predict(model)), pch = 19, 
     xlab="Measured Mean PPS", ylab="Modeled Mean PPS",
     xlim=c(0, 1500), ylim=c(0,1500))
lines(c(0, 1500), c(0, 1500))

```

* Using this model, I predicted the PPS values for each station and year

```{r, echo=FALSE, results="markup"}

df <- (expand.grid("YEAR" = unique(means$YEAR), 
                   "MOORING" = unique(means$MOORING)))
#df <- merge(df, CXY[,c(1,4)], by="MOORING")

df <- as.data.table(cbind(df, "Mean" = exp(predict(model, df))))

dcast(df, MOORING~YEAR) # Predicted values for all stations and years

```

* And calculated the predicted PPS values for them missing moorings, which allowed me to calculate the CV

```{r, echo=FALSE, results="markup"}

mean.filled <- rbind.data.frame(means, df[c(7, 19, 11),])


df.m <- mean.filled[, .(MEAN = mean(Mean)), by = .(YEAR)] # Average of pred. means

df.m 

CV <- round(sd(df.m$MEAN)/mean(df.m$MEAN), digits=2)
```

* The CV of the mean predicted PPS values across the three years is `r CV`

* To reach a target CV of 0.08 (i.e., ability to detect a -25% change over 10 yrs), N = `r round(11*(CV^2)/(0.08^2))` sensors would be required.

# Generating new data using the model estimates

* From the model, I generated new data from n arbitrary stations over y years

```{r}

# then generate PPS values according to mixed effects model (as below)

# generating new data using the model output
 
 simFun <- function(n, y, r){
   
   # n is the number of moorings
   # y is the number of years
   # r is the growth rate over the entire period
   
   # Generate single random intercept term 
   int <- summary(model)$coefficients[1]
   i.se <- summary(model)$coefficients[4]
   i <- rnorm(1, int, i.se/sqrt(30))
   
   # Generate random effects for n moorings
   #sig.b0 <- exp(unlist(model$modelStruct$reStruct))*model$sigma
  # b0 <- rnorm(n, 0, sig.b0) 
   
   # Generate random effects for y years
  # yr.eff <- fixef(model)[2]
  # yr.se <- summary(model)$tTable[4]
  # b1 <- rnorm(y, yr.eff, yr.se/sqrt(30))
   
   # Add noise to rate of change in population over time
   if (r == 0) {cum.r <- rep(1, y)} else {
   rvals <- rnorm(y-2, mean=((1+r)^(1/(y-1))), sd=0.05) # sd is arbitrary
   rvals[y-1] <- (1+r)/prod(rvals[1:(y-2)]) # solve for final decline
   rvals <- rvals[sample(1:(y-1), (y-1))] 
   cum.r <- c(1, cumprod(rvals))}
   
   sdev <- VarCorr(model)
   
   df <- expand.grid("MOORING" = 1:n, 
                     "YEAR" = 1:y, 
                     "PPS" = NA)
   
   # PPS = exp(intercept + mooring effect + noise) * rate change
   # removed: + b1[df$YEAR]
   df$PPS <- exp(i + b0[df$MOORING]  + rnorm(n*y, 0, sd=sdev)) * cum.r[df$YEAR] 
   
   return(df)
  
 }

new.data <- simFun(n = 50, y = 10, r = 0)

plot(new.data$YEAR, new.data$PPS)

new.model <- lme(fixed = log(PPS) ~ YEAR, 
             random = ~ 1 | MOORING, data = new.data)

summary(new.model)

plot(new.model, resid(.) ~ fitted(.))

```



```{r, SimSetup, results="markup"}

  
  # using the same mixed-effects model as above

n.moorings <- c(10, 50, 100)
p.change <- seq(-0.5, 0.5, by = 0.05)
nrep <- 1000
tab.res <- data.frame()

for (no in n.moorings){
  
  for (pc in p.change){
    
      r.est <- vector()
  
      for (i in 1:nrep){
    
        nd <- simFun(no, 10, pc)
    
        #nm <- lme(fixed = log(PPS) ~ YEAR, 
        #     random = ~ 1 | MOORING, data = nd)
        
        nm <- lm(log(PPS) ~ YEAR, data=nd)
    
        ye <- ifelse(summary(nm)$coefficients[8]<0.05, 1, 0)
    
        r.est <- c(r.est, ye)
    
      } # end i in n.rep
      
      tab.res <- rbind.data.frame(tab.res, 
                                  cbind.data.frame("No.M" = no, "P.Change" = pc, 
                                                   "P.Dect" = sum(r.est)/length(r.est)))
      
  } # end pc in p.change
    
    
  } # end nm in no. moorings


table <- dcast(tab.res, No.M ~ P.Change)

print(table)

```

```{r, results="markup"}

# Adding spatial component to the model

load("./Data/CXY.RData") # CXY
load("./Data/predGrid.RData") #gridOut


names(CXY)[4]<- "DENSITY"
names(CXY)[1] <- "MOORING"

means <-  PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(YEAR, MOORING)]

means <- merge(CXY[,c(1,4)], means, by="MOORING")
 
D.model <- glm(log(Mean) ~ as.numeric(factor(YEAR)) + log(DENSITY), data = means)

summary(D.model)

pred.grid <- expand.grid("DENSITY"=seq(0.1, 2.5, by=0.1), "YEAR"=1:3)

p.vals <- predict(D.model, pred.grid)

pred.grid <- cbind.data.frame(pred.grid, "Pred"=p.vals)

ggplot(pred.grid, aes(x=DENSITY, y=Pred, group=YEAR, color=YEAR))+
  geom_line()

```

```{r}

simSpatial <- function(n, y, r){
   
   # n is the number of moorings
   # y is the number of years
   # r is the growth rate over the entire period
  
   # Choose random X, Y locations for n moorings and get D vals
   #gO <- gridOut[which(gridOut$D>-40 & gridOut$D <=-10),]
   d.n <- log(gridOut$PpSqKm[sample(1:nrow(gridOut), n, replace=TRUE)])
   
   # Generate single random intercept term 
   int <- D.model$coefficients[1]
   i.se <- summary(D.model)$coefficients[4]
   i <- rnorm(1, int, i.se/sqrt(30))
   
   # Generate density effects for n spatial locs
   d.eff <- summary(D.model)$coefficients[3]
   d.se <- summary(D.model)$coefficients[6]
   b1 <- rnorm(n, d.eff, d.se/sqrt(30))
   
   # Add noise to rate of change in population over time
   if (r == 0) {cum.r <- rep(1, y)} else {
   rvals <- rnorm(y-2, mean=((1+r)^(1/(y-1))), sd=0.05) # sd is arbitrary
   rvals[y-1] <- (1+r)/prod(rvals[1:(y-2)]) # solve for final decline
   rvals <- rvals[sample(1:(y-1), (y-1))] 
   cum.r <- c(1, cumprod(rvals))}
   
   sdev <- sigma(D.model)
   
   df <- expand.grid("MOORING" = 1:n, 
                     "YEAR" = 1:y, 
                     "DENSITY" = NA, 
                     "PPS"= NA)
   df$DENSITY <- exp(rep(d.n, y))
   
   # PPS = exp(intercept + mooring effect + density effect + noise) * rate change
   # not included: + b1[df$YEAR]
   df$PPS <- exp(i + (b1*d.n)[df$MOORING] + rnorm(n*y, 0, sd=sdev)) * cum.r[df$YEAR] 
   
   return(df)
  
 }


new.data <- as.matrix(simSpatial(n = 50, y = 10, r = -.25))

new.data <- as.data.frame(new.data, names=c("MOORING", "YEAR", "DENSITY", "PPS"))

plot(new.data$YEAR, new.data$PPS)

new.model <- lme(fixed = log(PPS) ~ YEAR, 
             random = ~ 1 | MOORING, data = new.data)

summary(new.model)

plot(new.model, resid(.) ~ fitted(.))


```

```{r}

n.moorings <- c(10, 100)
p.change <- seq(-0.5, 0.5, by = 0.05)
nrep <- 1000
tab.res <- data.frame()

for (no in n.moorings){
  
  for (pc in p.change){
    
      r.est <- vector()
  
      for (i in 1:nrep){
    
        nd <- as.matrix(simSpatial(no, 10, pc))

        nd <- as.data.frame(nd, names=c("MOORING", "YEAR", "DENSITY", "PPS"))
    
        nm <- lme(fixed = log(PPS) ~ YEAR, 
             random = ~ 1 | MOORING, data = nd)
        
        ye <- ifelse(summary(nm)$tTable[10]<0.05 & sign(summary(nm)$tTable[2])==sign(pc), 1, 0)
    
        r.est <- c(r.est, ye)
    
      } # end i in n.rep
      
      tab.res <- rbind.data.frame(tab.res, 
                                  cbind.data.frame("No.M" = no, "P.Change" = pc, 
                                                   "P.Dect" = sum(r.est)/length(r.est)))
      
  } # end pc in p.change
    
    
  } # end nm in no. moorings

cbPalette <- c("#E69F00", "#56B4E9", "#009E73","#0072B2",  "#CC79A7")

ggplot()+
  geom_line(data=tab.res, aes(x=P.Change, y=P.Dect, 
                              group=as.factor(No.M), color=as.factor(No.M)), size=1.5)+
  scale_color_manual(name="No. Moorings", values=cbPalette)+
  geom_line(data=data.frame("P.Change"=c(-0.5, 0.5), "P.Dect"=c(0.8, 0.8)), 
            aes(x=P.Change, y=P.Dect), linetype=4, size=2)+
  geom_line(data=data.frame("P.Change"=c(-0.5, -0.5), "P.Dect"=c(0.14, 0.33)), 
            aes(x=P.Change, y=P.Dect) ,size =3, color = "red")+
  xlab("Percent Change in Population")+
  ylab("Power")+
  scale_x_continuous(labels=scales::percent)+
  theme_bw()+
  theme(legend.key=element_blank())

table <- dcast(tab.res, No.M ~ P.Change)

print(table)


```