
---
title: "Bayesian Power of PAM for MBay Pp"
header-includes:
   - \usepackage{caption}
output: 
  pdf_document:
    latex_engine: xelatex

---
\begin{center}
\Large{Eiren Jacobson} 
\end{center}

\begin{center}
\large{`r format(Sys.time(), '%d %B %Y')`}
\end{center}

```{r ChunkOpts, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, results="hide", message=FALSE, warning=FALSE, fig.align="center")
```

# Using MBay data to generate parameters for simulation

* The mean PPS values measured across all stations and years is shown in the following table

```{r, results="show", echo=FALSE}
 
 require(data.table)
 require(ggplot2)
 require(reshape2)
 setwd("~/Desktop/acSim/")
 
 load("./Data/CpodPPS.RData")
 
 mbay.data <-  PPS.Data[,.(PPS = mean(PPS)), 
                           by = .(YEAR, MOORING)]
 mean.table <-  dcast(mbay.data, MOORING~YEAR)

 
 mean.table$ALL <- PPS.Data[,.(PPS = mean(PPS)), 
                           by = .(MOORING)]$PPS

# d <- PPS.Data[,.(Mean = mean(PPS), Var = var(PPS)), 
#                            by = .(MOORING)]

mean.table <- rbind.data.frame(mean.table, 
                               cbind("MOORING"="ALL", "2013"=221.0256, "2014"=451.5736, "2015"=365.3147, "ALL" = 350.781))

mean.table[,2] <- as.numeric(mean.table[,2])
mean.table[,3] <- as.numeric(mean.table[,3])
mean.table[,4] <- as.numeric(mean.table[,4])
mean.table[,5] <- as.numeric(mean.table[,5])

mean.table[,2:5] <- round(mean.table[,2:5], digits=0)

mean.table

```

* First, I combined these passive acoustic data with porpoise density estimates 
* The density estimates come from a GAM of aerial survey data collected between 2000 and 2013

```{r, echo=FALSE, results="markup", height=2}

load("./Data/CXY.RData") # CXY

names(CXY)[1] <- "MOORING"

mbay.data <- as.data.table(merge(CXY, mbay.data, by="MOORING"))

m <- mbay.data[,.(Mean = mean(PPS), Density=mean(Density)), 
                           by = .(MOORING)]

plot(log(m$Density), log(m$Mean), pch=19,
     xlab = "log(Mean Porpoise Density)", ylab="log(Mean PPS)", main = "Real Data")
```

```{r, results="markup", echo=FALSE, width=3, height=3}


load("./Data/predGrid.RData") #gridOut
load("./Data/coastXY.RData") # coast

ggplot()+
  geom_polygon(data=coast, aes(x=X/1000, y=Y/1000), fill="gray")+
  geom_point(data=gridOut, aes(x=X/1000, y=Y/1000, color=PpSqKm), size=1)+
  scale_color_gradient2(name=bquote('Density ('*km^-2*')'),
                        low="dodgerblue2", mid="yellow", high="red", 
                        midpoint=max(gridOut$PpSqKm)/2, limits=c(0, max(gridOut$PpSqKm)))+
  theme_bw()+
  coord_equal(xlim=c(-50, 50), ylim=c(-50,50))+
  xlab("X (km)")+
  ylab("Y (km)")+
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(),
        legend.key=element_blank(),
        legend.title=element_text(face="bold", size=10),
        
        strip.background=element_blank(),
        panel.margin=unit(1.25, "lines"),
        legend.key.height=unit(c(1.25), "lines"),
        plot.margin=unit(c(0, 0, 0, 0), "lines"),
        legend.position=c(0.2, 0.3))



```

* Now, to build a Bayesian model 

```{r}

require(R2jags)

model.string <- "
    
  model{
  
  # Priors
  # Parameters to be estimated
  beta1 ~ dunif(0, 2) # rate of change between years
  beta2 ~ dunif(0, 1000) # effect of density
  a ~ dunif(0.001, 10) # shape/scale parameter for gamma

  # Likelihood
  for (i in 1:nsite){

      # Underlying 'true' mean density from GAM/spline estimate
      d[i] ~ dlnorm(D.mu[i], D.tau[i]) # D.mu & D.tau are l-transformed
    
    for (j in 1:nyear){
    
    C[i, j] ~ dpois(n.hat[i, j]*rho[i, j]) # NB distribution
    rho[i, j] ~ dgamma(a, a) 
    n.hat[i, j] <- pow(beta1,j)*(beta2*d[i])  # predictor
    
    } # end for j
    
  } # end for i

  } # end model

"

x <- dcast(mbay.data, MOORING ~ YEAR, value.var="PPS")

data <- list(C = round(x[,-1]), nsite=length(unique(mbay.data$MOORING)), 
             nyear = length(unique(mbay.data$YEAR)),
             D.mu = log(unique(mbay.data$Density)),
             D.tau = 1/log(unique(mbay.data$SE.D)^2/unique(mbay.data$Density)^2 + 1))

inits <- list(beta1 = 1,
              beta2 = 1,
              a = 1)

model.out <- jags.model(file=textConnection(model.string),
                          data, 
                          inits = inits,
                          n.adapt = 1000, 
                          n.chains = 4, 
                          quiet = T)

samples <- coda.samples(model.out, c('beta1', 'beta2'), n.iter=10000, thin=10)

pars <- summary(samples)$statistics

model <- glm(round(log(mbay.data$PPS), digits=0) ~ y + log(mbay.data$Density), family=poisson, link=log)

eg <- expand.grid("y"=1:3, "d"=seq(0, 3, by=0.25), "pps" = NA)
eg$pps <- (pars[1]^eg$y)*(pars[2]*eg$d)

names(eg)[1] <- "YEAR"
ggplot()+
  geom_point(data=mbay.data, aes(x=Density, y=PPS, 
                                 color=as.factor(as.numeric(as.factor(mbay.data$YEAR)))))+
  geom_line(data=eg, aes(x=d, y=pps, color=as.factor(YEAR)))+
  scale_color_discrete(name="Year")+
  theme_bw()

```
