---
title: "Power of PAM for MBay Harbor Porpoise"
header-includes:
   - \usepackage{caption}
output: 
  pdf_document:
    latex_engine: xelatex

---
\begin{center}
\Large{Eiren Jacobson} 
\end{center}

\begin{center}
\large{`r format(Sys.time(), '%d %B %Y')`}
\end{center}

```{r ChunkOpts, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, results="hide", message=FALSE, warning=FALSE, fig.align="center")
```

# Introduction and Objectives

The ability to detect trends in cetacean abundance is crucial for effective management, particularly when populations may be declining due to cumulative sub-lethal anthropogenic impacts rather than mortality in fishing sets or other direct take.   Statistical power to detect trends in abundance with traditional survey data is poor.  For example, in 78% of dolphin and porpoise populations, a 50% decline would not be detected using current visual survey methods (Taylor et al., 2007).  Fixed passive acoustic monitoring is a promising method for monitoring cetacean populations (Marques et al., 2013).  Designing passive acoustic schemes for long-term monitoring of cetacean populations is now feasible; however, there is no framework for optimizing passive acoustic surveys to maximize their precision and the resulting statistical power to detect trends in abundance.  

I will use the Monterey Bay population of harbor porpoise as a case study to develop design criteria for passive acoustic monitoring networks for cetaceans. Between 2000 and 2013, over 100 fine-scale aerial line-transect surveys were flown in the Monterey Bay region.  In 2013, 2014, and 2015, a grid of 11 passive acoustic monitoring instruments (C-PODs; Chelonia Ltd., http://chelonia.co.uk) was deployed in the Monterey Bay study area.  I will use these aerial survey and passive acoustic data collected from the Monterey Bay population of harbor porpoise to create simulated datasets representing hypothetical monitoring network designs and changes in abundance.  From these simulations I will quantify the expected statistical power of passive acoustic surveys to detect trends in cetacean abundance.

**Some Questions:**

1. What precision is required to detect trends in abundance of different magnitudes, and what is the natural variability of the population?
2. How variable is the passive acoustic dataset across the three years of C-POD deployments?
3. How does spatial variability influence power to detect trends?
4. In a frequentist framework, how does the power of PAM to detect trends in abundance compare to aerial surveys?
5. What would a Bayesian framework to optimize PAM look like?
6. How would optimal PAM design differ for goals of detecting long-term trends vs. short-term disturbance?

\newpage

# 1. Estimating natural variability w.r.t. target CVs for monitoring

## Using TRENDS to determine target CV

Following Rojas-Bracho et al. (2010), I used the program TRENDS (Gerrodette 1987) to determine the required precision to detect trends:

\begin{table}[h!]
\centering
\caption{Required precision (CVs) to detect changes of varying magnitudes.  Power = 0.8, Alpha = 0.05, Test = two-tailed.}
\label{tab:echo}
\def\arraystretch{1.25}

\begin{tabular}{ccc}

\vspace{3pt}
\textbf{Change in Pop.} & 
\textbf{No. Yrs} &
\textbf{Target CV}\\
\hline
-50\% & 10 & 0.18 \\
-25\% & 10 & 0.08 \\
-10\% & 10 & 0.03 \\
-5\% & 5 & 0.02 \\
\hline

\end{tabular}

\end{table}

## A simple population model to estimate natural variability

Next, I used a population model to determine the expected variance in the abundance of the population given random processes of birth and death.  The code for this is in the file popModel.R.  In summary, using life history parameters from Taylor et al. (2007) as a guide, I constructed a simple model of the Monterey Bay harbor porpoise population. 

Assumptions:

* N = 2,500 = K (based on Forney et al. 2014, where the lower 20th percentile of the abundance estimate was 2,480)
* Calf survival (1st year) S0 = 0.789, juvenile and adult survival S1 = 0.95
* Individuals reach reproductive maturity at age 6
* Taylor et al. (2007) report that the inter-birth interval = 1 year; however, a population at carrying capacity with the survivorship parameters described above and no other sources of mortality must have a fecundity lower than 0.5.  I therefore set the maximum fecundity = 0.5 (i.e., all adult females produce a calf), minimum fecundity = 0.165 (trial and error . . . ) and allowed f in a given year to vary depending on N/K: `f <- fmin+((fmax-fmin)*(1-(sum(N[t,])/K)^z))` with a compensation coefficient z = 3.
* The initial age structure of the population was set so that 50% of individuals were adults, and the remainder of the population was allocated amongst the other age classes according to the survivorship expectations (a.k.a math)

By running this population forward and experimenting with different values of fmin and fmax, I determined that at carrying capacity, f = 0.165 produces a stable harbor porpoise population over a 100 year period. 

Next, I used the values of S0, S1, and f to simulate the change in a population from YR1 to YR2 due to stochasticity.

For each individual in each age class, I simulated survival/non-survival, male/female, and calf/no calf as Bernoulli trials with probabilities given by S0 = 0.789, S1 = 0.95, and f = 0.165, with a starting population N = 2,500.

```{r, caption="Population size in year 2 given a starting population of N = 2500 and random birth and death processes."}

N.vec <- c(150, 115, 110, 105, 100, 95, 1825) # initial population structure 
N.sto <- rep(NA, 10000) # number of iterations
S0 = 0.789 # calf survival
S1 = 0.95 # adult survival
f = 0.165 # fecundity

f <- 0.165
for (i in 1:10000){
  
  nA <- sum(rbinom(N.vec[6], 1, S1)) + sum(rbinom(N.vec[7], 1, S1)) 
  n0 <- sum(rbinom(sum(rbinom(N.vec[7], 1, 0.5)), 1, f))
  n1 <- sum(rbinom(N.vec[1], 1, S0))
  n2 <- sum(rbinom(N.vec[2], 1, S1))
  n3 <- sum(rbinom(N.vec[3], 1, S1))
  n4 <- sum(rbinom(N.vec[4], 1, S1))
  n5 <- sum(rbinom(N.vec[5], 1, S1))
  
  N.sto[i] <- sum(nA + n0 + n1 + n2 + n3 + n4 + n5)
  
  }
  
 hist(N.sto, col="gray", main = "", xlab="Abundance in Year 2")

```

The variation in the population due to random chance leads to a CV of `r round(sd(N.sto)/mean(N.sto), digits = 3)`, which is an order of magnitude lower than the smallest target CVs shown in Table 1.  Therefore, I conclude that in a population of this size, stochasticity in birth and death processes produce changes in the population that are small relative to the population size.  I do not expect natural variability to limit statistical power to monitor the population.

# 2. Variability in the 3-year passive acoustic dataset

* The mean PPS values measured across all stations and years is shown in the following table

```{r, results="show", echo=FALSE}
 
 require(data.table)
 require(ggplot2)
 require(reshape2)
 require(nlme)
 
 setwd("~/Desktop/acSim/")
 load("./Data/CpodPPS.RData")
 
 means <-  PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(YEAR, MOORING)]
 mean.table <-  dcast(means, MOORING~YEAR)

 
 mean.table$ALL <- PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(MOORING)]$Mean

d <- PPS.Data[,.(Mean = mean(PPS), Var = var(PPS)), 
                           by = .(MOORING)]

mean.table <- rbind.data.frame(mean.table, 
                               cbind("MOORING"="ALL", "2013"=221.0256, "2014"=451.5736, "2015"=365.3147, "ALL" = 350.781))

mean.table[,2] <- as.numeric(mean.table[,2])
mean.table[,3] <- as.numeric(mean.table[,3])
mean.table[,4] <- as.numeric(mean.table[,4])
mean.table[,5] <- as.numeric(mean.table[,5])

mean.table[,2:5] <- round(mean.table[,2:5], digits=0)

mean.table

```

* I first constructed a linear mixed-effects model where year is a fixed effect and mooring is a random effect.  The response variable is the log(mean PPS) for each mooring and year.

```{r, echo=TRUE, results="markup"}
 
model <- lme(fixed = log(Mean) ~ as.numeric(factor(YEAR)), 
             random = ~ 1 | MOORING, data = means)

summary(model)

```


```{r, echo=FALSE}

plot(means$Mean, exp(predict(model)), pch = 19, 
     xlab="Measured Mean PPS", ylab="Modeled Mean PPS",
     xlim=c(0, 1500), ylim=c(0,1500))
lines(c(0, 1500), c(0, 1500))

```

* Using this model, I predicted the PPS values for each station and year

```{r, echo=FALSE, results="markup"}

df <- (expand.grid("YEAR" = unique(means$YEAR), 
                   "MOORING" = unique(means$MOORING)))
#df <- merge(df, CXY[,c(1,4)], by="MOORING")

df <- as.data.table(cbind(df, "Mean" = exp(predict(model, df))))

dcast(df, MOORING~YEAR) # Predicted values for all stations and years

```

* And calculated the predicted PPS values for the missing moorings, which allowed me to calculate the CV

```{r, echo=FALSE, results="markup"}

mean.filled <- rbind.data.frame(means, df[c(7, 19, 11),])


df.m <- mean.filled[, .(MEAN = mean(Mean)), by = .(YEAR)] # Average of pred. means

df.m 

CV <- round(sd(df.m$MEAN)/mean(df.m$MEAN), digits=2)
```

* The CV of the mean predicted PPS values across the three years is `r CV`

* To reach a target CV of 0.08 (i.e., ability to detect a -25% change over 10 yrs), N = `r round(11*(CV^2)/(0.08^2))` sensors would be required.

## Generating new data using the model estimates

From the model, I generated new data from n arbitrary stations over y years

```{r, echo=TRUE, eval=FALSE}

# generating new data using the model output
 
 simFun <- function(n, y, r){
   
   # n is the number of moorings
   # y is the number of years
   # r is the growth rate over the entire period
   
   # Generate single random intercept term 
   int <- summary(model)$coefficients$fixed[1]
   i.se <- summary(model)$tTable[3]
   i <- rnorm(1, int, i.se/sqrt(30))
   
   # Generate random effects for n moorings
   sig.b0 <- exp(unlist(model$modelStruct$reStruct))*model$sigma
   b0 <- rnorm(n, 0, sig.b0) 
   
   # Generate random effects for y years
   yr.eff <- fixef(model)[2]
   yr.se <- summary(model)$tTable[4]
   b1 <- rnorm(y, yr.eff, yr.se/sqrt(30))
   
   # Add noise to rate of change in population over time
   if (r == 0) {cum.r <- rep(1, y)} else {
   rvals <- rnorm(y-2, mean=((1+r)^(1/(y-1))), sd=0.05) # sd is arbitrary
   rvals[y-1] <- (1+r)/prod(rvals[1:(y-2)]) # solve for final decline
   rvals <- rvals[sample(1:(y-1), (y-1))] 
   cum.r <- c(1, cumprod(rvals))}
   
   sdev <- as.numeric(VarCorr(model)[4])
   
   df <- expand.grid("MOORING" = 1:n, 
                     "YEAR" = 1:y, 
                     "PPS" = NA)
   
   # PPS = exp(intercept + mooring effect + noise) * rate change
   # removed: + b1[df$YEAR]
   df$PPS <- exp(i + b0[df$MOORING]  + rnorm(n*y, 0, sd=sdev)) * cum.r[df$YEAR] 
   
   return(df)
  
 }

# checking that this simulation works (results not shown)

new.data <- simFun(n = 50, y = 10, r = 0)

plot(new.data$YEAR, new.data$PPS)

new.model <- lme(fixed = log(PPS) ~ YEAR, 
             random = ~ 1 | MOORING, data = new.data)

summary(new.model)

plot(new.model, resid(.) ~ fitted(.))

```

Then, using this data generation code, I simulated different possible scenarios:

```{r, SimSetup, results="markup", echo=FALSE, eval=FALSE}

# Using the data generation code above, 

n.moorings <- c(10, 25, 50, 75, 100)
p.change <- c(-0.5, -0.25, -.1, 0, 1.25)
nrep <- 1000
tab.res <- data.frame()

for (no in n.moorings){
  
  for (pc in p.change){
    
      r.est <- vector()
  
      for (i in 1:nrep){
    
        nd <- simFun(no, 10, pc)
    
        nm <- lme(fixed = log(PPS) ~ YEAR, 
              random = ~ 1 | MOORING, data = nd)
    
        ye <- ifelse(summary(nm)$tTable[10]<0.05, 1, 0)
    
        r.est <- c(r.est, ye)
    
      } # end i in n.rep
      
      tab.res <- rbind.data.frame(tab.res, 
                                  cbind.data.frame("No.M" = no, "P.Change" = pc, 
                                                   "P.Dect" = sum(r.est)/length(r.est)))
      
    } # end pc in p.change
    
  } # end no in no. moorings

 table <- dcast(table, No.M ~ P.Change, value.var = "P.Dect")
 
 save(table, file="./Data/nonSpatialPower.RData")
 
```

```{r, results="markup"}
 
load("./Data/nonSpatialPower.RData")

print(round(table, digits=2))
 
```

# 3. Spatial variability

To add spatial variability, I used the 10 years of aerial survey data to estimatea an average density surface across the study area, then modelled PPS as a function of year + underlying mean density.  This allowed me to generate new data for any point in the study area.

\includegraphics[width=250pt]{~/Desktop/acSim/GAMOut.png}

```{r, results="markup", eval=FALSE}

# Adding spatial component to the model

load("./Data/CXY.RData") # CXY
load("./Data/predGrid.RData") #gridOut


names(CXY)[4]<- "DENSITY"
names(CXY)[1] <- "MOORING"

means <-  PPS.Data[,.(Mean = mean(PPS)), 
                           by = .(YEAR, MOORING)]

means <- merge(CXY[,c(1,4)], means, by="MOORING")
 
D.model <- glm(log(Mean) ~ as.numeric(factor(YEAR)) + log(DENSITY), data = means)

summary(D.model)

pred.grid <- expand.grid("DENSITY"=seq(0.1, 2.5, by=0.1), "YEAR"=1:3)

p.vals <- predict(D.model, pred.grid)

pred.grid <- cbind.data.frame(pred.grid, "Pred"=p.vals)

ggplot(pred.grid, aes(x=DENSITY, y=Pred, group=YEAR, color=YEAR))+
  geom_line()

```

```{r, echo=TRUE, eval=FALSE}

simSpatial <- function(n, y, r){
   
   # n is the number of moorings
   # y is the number of years
   # r is the growth rate over the entire period
  
   # Choose random X, Y locations for n moorings and get D vals
   #gO <- gridOut[which(gridOut$D>-40 & gridOut$D <=-10),]
   d.n <- log(gridOut$PpSqKm[sample(1:nrow(gridOut), n, replace=TRUE)])
   
   # Generate single random intercept term 
   int <- D.model$coefficients[1]
   i.se <- summary(D.model)$coefficients[4]
   i <- rnorm(1, int, i.se/sqrt(30))
   
   # Generate density effects for n spatial locs
   d.eff <- summary(D.model)$coefficients[3]
   d.se <- summary(D.model)$coefficients[6]
   b1 <- rnorm(n, d.eff, d.se/sqrt(30))
   
   # Add noise to rate of change in population over time
   if (r == 0) {cum.r <- rep(1, y)} else {
   rvals <- rnorm(y-2, mean=((1+r)^(1/(y-1))), sd=0.05) # sd is arbitrary
   rvals[y-1] <- (1+r)/prod(rvals[1:(y-2)]) # solve for final decline
   rvals <- rvals[sample(1:(y-1), (y-1))] 
   cum.r <- c(1, cumprod(rvals))}
   
   sdev <- sigma(D.model)
   
   df <- expand.grid("MOORING" = 1:n, 
                     "YEAR" = 1:y, 
                     "DENSITY" = NA, 
                     "PPS"= NA)
   df$DENSITY <- exp(rep(d.n, y))
   
   # PPS = exp(intercept + mooring effect + density effect + noise) * rate change
   # not included: + b1[df$YEAR]
   df$PPS <- exp(i + (b1*d.n)[df$MOORING] + rnorm(n*y, 0, sd=sdev)) * cum.r[df$YEAR] 
   
   return(df)
  
 }


```

```{r, echo=FALSE, eval=FALSE}

n.moorings <- c(10, 100)
p.change <- seq(-0.5, 0.5, by = 0.05)
nrep <- 1000
tab.res <- data.frame()

for (no in n.moorings){
  
  for (pc in p.change){
    
      r.est <- vector()
  
      for (i in 1:nrep){
    
        nd <- as.matrix(simSpatial(no, 10, pc))

        nd <- as.data.frame(nd, names=c("MOORING", "YEAR", "DENSITY", "PPS"))
    
        nm <- lme(fixed = log(PPS) ~ YEAR, 
             random = ~ 1 | MOORING, data = nd)
        
        ye <- ifelse(summary(nm)$tTable[10]<0.05 & sign(summary(nm)$tTable[2])==sign(pc), 1, 0)
    
        r.est <- c(r.est, ye)
    
      } # end i in n.rep
      
      tab.res <- rbind.data.frame(tab.res, 
                                  cbind.data.frame("No.M" = no, "P.Change" = pc, 
                                                   "P.Dect" = sum(r.est)/length(r.est)))
      
  } # end pc in p.change
    
    
  } # end nm in no. moorings

cbPalette <- c("#E69F00", "#56B4E9", "#009E73","#0072B2",  "#CC79A7")

ggplot()+
  geom_line(data=tab.res, aes(x=P.Change, y=P.Dect, 
                              group=as.factor(No.M), color=as.factor(No.M)), size=1.5)+
  scale_color_manual(name="No. Moorings", values=cbPalette)+
  geom_line(data=data.frame("P.Change"=c(-0.5, 0.5), "P.Dect"=c(0.8, 0.8)), 
            aes(x=P.Change, y=P.Dect), linetype=4, size=2)+
  geom_line(data=data.frame("P.Change"=c(-0.5, -0.5), "P.Dect"=c(0.14, 0.33)), 
            aes(x=P.Change, y=P.Dect) ,size =3, color = "red")+
  xlab("Percent Change in Population")+
  ylab("Power")+
  scale_x_continuous(labels=scales::percent)+
  theme_bw()+
  theme(legend.key=element_blank())

table <- dcast(tab.res, No.M ~ P.Change)

print(table)


```

# 4. How does the power of PAM to detect trends in abundance compare to aerial surveys?

\includegraphics[width=400pt]{~/Desktop/acSim/acPwr.png}

# 5. Gettin' Bayesian

Instead of using the mixed-effects or GLM models described above, I'd like to use a Bayesian model to estimate the effects of underlying density and year/trend on the population, and use the posterior estimates to generate new datasets. 

```{r, echo=TRUE}

require(R2jags)

model.string <- "
    
  model{
  
  # Priors
  # Parameters to be estimated
  beta1 ~ dunif(0, 2) # rate of change between years
  beta2 ~ dunif(0, 1000) # effect of density
  a ~ dunif(0.001, 10) # shape/scale parameter for gamma

  # Likelihood
  for (i in 1:nsite){

    # Underlying 'true' mean density from GAM/spline estimate
    d[i] ~ dlnorm(D.mu[i], D.tau[i]) # D.mu & D.tau are l-transformed
    
    for (j in 1:nyear){
    
    C[i, j] ~ dpois(n.hat[i, j]*rho[i, j]) # NB distribution
    rho[i, j] ~ dgamma(a, a) 
    n.hat[i, j] <- pow(beta1,j)*(beta2*d[i])  # predictor
    
    } # end for j
    
  } # end for i

  } # end model

"
```




